Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-11-12 paper/paper.tex
intro minor changes

\section{Introduction}
\label{sec:intro}
\steven{This intro sounds reallchoppy but follows the following outline.

  \begin{itemize}
  \item parsing correctness is imporant because parsing is
  everywhere and failure
          could even invalidate other proof efforts.
  \item verified parsers should be
          easier to write
  \item Conal Elliott gives the right idea for what the foundation
          of parser verification should be --- type theory

  \end{itemize}
          }
Formal verification encompasses a broad suite of techniques one may use to
mathematically guarantee the correctness of software systems. We can only
reap the benefits of
any proof effort if its underlying assumptions hold. For instance, verifying the
functional correctness of an algorithm is no use if it is run on buggy hardware.
Likewise, the promised guarantees of a program may be jeopardized if it is
parsed incorrectly.

Large verified developments, like CompCert \cite{leroy_formal_2009},
invest enormous amounts of person-hours into proof development. In early
versions of CompCert, the apparent
assurances associated with their effort could indeed be invalidated through
parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
throughout computing and yet parsers are frequently unverified, even in
otherwise verified systems. The CompCert team later
developed a method for validating the correctness of LR parsers
\cite{jourdanValidatingLRParsers2012} and then replaced their parser with one
that is certified to be correct. Despite their success in  validating their C parser and other work in the
generation of  verified parsers
\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
it is still arduous and highly nontrivial to write and verify a
parser. Further, distinct approaches to generating verified parsing
implementations are often relative to a different grammar formalism. We believe
that these sepearate grammar formalisms can all be unified into a
syntax-independent notion of grammar, and that this notion of grammar can be
axiomatized into a syntax that is useful for building verified parsers.

\steven{I don't like how I make the cited work sound bad above}

Elliott \cite{elliottSymbolicAutomaticDifferentiation2021}
notes that grammars can be described as type-level predicates on strings that prove
language membership. That is, a \emph{formal grammar} $A$ is a function
$\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
showing that $w$ belongs to the language recognized by $A$. We adopt this
definition of a formal grammar.

In this paper, we present \theoryname (\theoryabbv) as a domain specific
type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
theory: the linear types can be interpreted as grammars, whereas the
non-linear types can be interpreted as ordinary sets. As in prior work
on dependent linear-non-linear types, the linear types are allowed to
depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.

@@

\paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
by giving small examples of programs in \theoryabbv so that we may build intuition.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. In \cref{sec:applicatoins} we demonstrate the
applicablity of \theoryabbv for defining parser terms and representing results
from formal language theory as proof terms in our calculus.
Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
that makes precise the connection between \theoryabbv syntax and formal grammars.

Our contributions are then:
%
\begin{itemize}
  \item \theoryname (\theoryabbv): A dependent
        linear-non-linear type theory for building verified parsers
  \item An implementation of \theoryabbv, which is embedded shallowly in Cubical
        Agda. Throughout this paper, we will use \agdalogo~ to mark results that
        have been mechanized in our Agda development.
\end{itemize}

@@

In \cref{litNFATterms} we define a family of grammars $A$ and then prove that each $A~s$ is
equivalent to the accepting traces in $N_{c}$ out of state $s$. Then the desired
equivalence arises from taking $s = 0$.

@@

Note that each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
identity on each $A~s$. To complete the proof of strong equivalence it remains
to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
$Trace~s~true$, which we show by using the equalizer type.

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
\newcommand{\balanced}{\mathtt{balanced}}
\newcommand{\Dyck}{\mathtt{Dyck}}
Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
grammars (CFGs). The smallest nontrivial CFG is the Dyck grammar
$\Dyck$ --- the grammar of balanced parentheses.

\pedro{Is it actually the smallest? How do you quantify size of nontrivial things?
  Might be better to just say ``an interesting nontrivial CFG [...]''}

@@

  Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
  $A_1$ to $A_2$ is a function assigning to each string $w$ a function
  $f_w : A_1(w) \to A_2(w)$.
\end{definition}
\pa{Do we actually need a definition of formal language morphism?
In case we do, it should probably be its own definition.}
First, note that parse transformers generalize language inclusion: if
$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
In our denotational semantics, linear terms will be interpreted as
such parse transformers, and the notions of unambiguous grammar,
parseability, etc, introduced in \Cref{sec:concepts} can be verified
to correspond to their intended meanings under this interpretation.

>>>>

\section{Introduction}
\label{sec:intro}
Formal verification encompasses a broad suite of techniques one may use to
mathematically guarantee the correctness of software systems. We can only
reap the benefits of
any proof effort if its underlying assumptions hold. For instance, verifying the
functional correctness of an algorithm is no use if it is run on buggy hardware.
Likewise, the promised guarantees of a program may be jeopardized if it is
parsed incorrectly.

Large verified developments, like CompCert \cite{leroy_formal_2009},
invest a significant number of person-hours into proof development. In early
versions of CompCert, the apparent
assurances associated with their effort were at risk of invalidation by
parser bugs \cite{yangFindingUnderstandingBugs}. Parsing is a ubiquitous task
throughout computing and yet parsers are frequently unverified, even in
otherwise verified systems. The CompCert team later
developed a method for validating the correctness of LR parsers
\cite{jourdanValidatingLRParsers2012} and then replaced their parser with one
that is certified to be correct. Despite their success in  validating their C parser and other work in the
generation of  verified parsers
\cite{lasserCoStarVerifiedALL2021,firsovCertifiedNormalizationContextFree2015,danielssonTotalParserCombinators2010},
it is still arduous and highly nontrivial to write a verified
parser. Further, approaches are sensitive to a choice of grammar and automata formalism to serve
as a
foundation, which can make it difficult to relate and interoperate different
styles of verified parsers.

In \cite{elliottSymbolicAutomaticDifferentiation2021}, Elliott describes formal
grammar as type-level predicates on strings that prove
language membership. That is, a \emph{formal grammar} $A$ is a function
$\String \to \Set$ such that for a string $w$, $A~w$ is the set of proofs
showing that $w$ belongs to the language recognized by $A$. This typed approach
to parsing serves a general setting that can simultaneuously interpret many distinct
grammar formalisms. In this paper, we adopt Elliott's view on formal grammars and present a
syntax that axiomatizes this semantic notion of grammar.

We present \theoryname (\theoryabbv) as a domain specific
type theory for building verified parsers. The type theory itself is a form of dependent linear-non-linear type
theory: the linear types can be interpreted as grammars, whereas the
non-linear types can be interpreted as ordinary sets. As in prior work
on dependent linear-non-linear types, the linear types are allowed to
depend on the non-linear types but not vice-versa \cite{krishnaswami_integrating_2015}.

@@

Parsers written in \theoryabbv take on a linear functional style, which makes
them familiar to write and amenable to compositional verification techniques. We
implement \theoryabbv as a shallowly embedded domain-specific langauge in the
Cubical Agda proof assistant. Throughout this paper, we will use \agdalogo~ to
mark results that are mechanized in our Agda development.

\paragraph{Contributions} This paper begins in \cref{sec:type-theory-examples}
by studying small example programs from \theoryabbv to build intuition.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type theory
that axiomatizes the underlying structures needed to conduct formal grammar
theory. Later in \cref{sec:applications} we demonstrate the
applicability of \theoryabbv for relating familiar grammar and automata
formalisms as well as building concrete parsers.
Then in \cref{sec:semantics-and-metatheory}, we give a denotational semantics
that makes precise the connection between \theoryabbv syntax and formal grammars.

Our contributions are then:
%
\begin{itemize}
  \item \theoryname (\theoryabbv): A dependent
        linear-non-linear type theory for building verified parsers
  \item A denotational grammars semantics for \theoryabbv
  \item An implementation of \theoryabbv in Agda
\end{itemize}

@@

In \cref{fig:litNFAterms} we define a family of grammars $A$ that serve as
interpretations for each state in $N_{c}$. We then prove that each $A~s$ is
equivalent to the accepting traces through $N_{c}$ out of state $s$.

@@

Each $A~s$ is unambiguous, so $(fromN_{c}~s) \circ (toN_{c}~s)$ is equal to the
identity on each $A~s$. To complete the proof of strong equivalence it remains
to show that $(toN_{c}~s) \circ (fromN_{c}~s)$ is equal to the identity on
$Trace~s~true$, which we show by using the equalizer type.

@@

\subsection{The Dyck Grammar}
%% 3. Dyck language
\newcommand{\balanced}{\mathtt{balanced}}
\newcommand{\Dyck}{\mathtt{Dyck}}
Moving up the Chomsky hierarchy from regular grammars, we arrive at context-free
grammars (CFGs). An interesting nontrivial CFG is the Dyck grammar
$\Dyck$ --- the grammar of balanced parentheses.

@@

  Let $A_1,A_2$ be formal grammars. Then a parse transformer $f$ from
  $A_1$ to $A_2$ is a function assigning to each string $w$ a function
  $f_w : A_1(w) \to A_2(w)$.
\end{definition}
\pedro{Do we actually need a definition of formal language morphism?
In case we do, it should probably be its own definition.}
First, note that parse transformers generalize language inclusion: if
$A_1(w), A_2(w)$ are all subsets of a one-element set, then a parse
transformer is equivalent to showing that $A_1(w) \subseteq A_2(w)$.
In our denotational semantics, linear terms will be interpreted as
such parse transformers, and the notions of unambiguous grammar,
parseability, etc, introduced in \Cref{sec:concepts} can be verified
to correspond to their intended meanings under this interpretation.
